---
title: "Data structure"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

```{r myfunctions}
#| include: false

# Function returning column classes DF
var_class_2 <- function(DT) {
  variable_class <- sapply(DT, class)
  x <- as.data.frame(variable_class)
  x$var_name <- row.names(x)
  setDT(x)
  setnames(x,
    old = c("var_name", "variable_class"),
    new = c("Variable", "Class")
  )
  x <- x[, .(Variable, Class)]
  setDF(x)
  return(x)
}

set.seed(20260514)
```

We examine the structure of the four data tables in midfielddata: number of observations, number and class of variables, representative values, and database keys. 

```{r}
#| echo: false

library(gt)
wrapr::build_frame(
  "Table", "Each row is", "N students", "N rows", "N columns", "Mb memory" |
    "student", "a student",
  "97,555", "97,555", 13L, 17.3 |
    "term", "a student in a term",
  "97,555", "639,915", 13L, 72.8 |
    "course", "a student in a course in a term",
  "97,555", "3,289,532", 12L, 324.3 |
    "degree", "a graduate from a program in a term",
  "49,543", "49,665", 5L, 5.2
) |>
  gt() |>
  # tab_caption("Table 1. Practice data tables in midfielddata") |>
  tab_options(table.font.size = "small") |>
  opt_stylize(style = 1, color = "gray") |>
  tab_style(
    style = list(cell_fill(color = "#c7eae5")),
    locations = cells_column_labels(columns = everything())
  )
```

## Load data

*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfielddata)
library(data.table)
library(wrapr)
```


*Load data tables.*  &nbsp; Data tables can be loaded individually or collectively as needed. 

```{r}
# Load one table as needed
data(student)

# Or load multiple tables
data(course, term, degree)
```

 
 
 

## Variables

We list the variables in each dataset to provide an overview of the information available. Keys and composite keys in each data table are indicated. The student ID variable `mcid` is the primary key linking the four tables. 

```{r}
#| echo: false
#| comment: ""
cat("student")
x <- var_class_2(student)
x$Keys <- ""
sel <- x$Variable %chin% c("mcid")
x$Keys[sel] <- "key"
x

cat("\nterm")
x <- var_class_2(term)
x$Keys <- ""
sel <- x$Variable %chin% c("mcid", "term")
x$Keys[sel] <- "key"
x

cat("\ncourse")
x <- var_class_2(course)
x$Keys <- ""
sel <- x$Variable %chin% c("mcid", "term_course", "abbrev", "number")
x$Keys[sel] <- "key"
x

cat("\ndegree")
x <- var_class_2(degree)
x$Keys <- ""
sel <- x$Variable %chin% c("mcid", "cip6", "term_degree")
x$Keys[sel] <- "key"
x
```





## `student`

Student-level demographic information for `r length(unique(student$mcid))` degree-seeking undergraduate students. Data at the "student-level" refers to information collected by undergraduate institutions about individual students  at matriculation, for example, age, sex, and race/ethnicity.

The `student` data table has `r nrow(student)` observations of `r ncol(student)` variables structured in row-record form, that is, information associated with a particular ID occupies a single row---one record per student. The data dictionary is documented at `?student`.

```{r}
student
```

Student IDs and institution names have been anonymized to remove identifiable information. 

```{r}
# Anonymized ID sample
sample(student$mcid, 8)

# Anonymized institutions
sort(unique(student$institution))
```

Counts in each category. Race/ethnicity and sex are often used as grouping variables. 

```{r}
#| echo: false
options(datatable.print.nrows = 15)
```

```{r}
# N by institution
student[, .N, by = "institution"][order(-N)]

# N by race/ethnicity
student[, .N, by = "race"][order(-N)]

# N by sex
student[, .N, by = "sex"][order(-N)]

# N by race/ethnicity and sex
student[, .N, by = c("race", "sex")][order(-N)]
```

```{r}
#| echo: false
options(datatable.print.nrows = 3)
```











## `term`

Student-level term information for `r length(unique(term$mcid))` undergraduates including program code, academic standing, and grade point average for individual students in each term.

The `term` data table has `r nrow(term)` observations of `r ncol(term)` variables structured in block-record form, that is, records associated with a particular ID can span multiple rows---one record per term for every student. The data dictionary is documented at `?term`.

```{r}
term
```

Terms are encoded `YYYYT`, where 

- `YYYY` is the year at the start of the academic year, and  
- `T` encodes the semester or quarter---Fall (`1`), Winter (`2`), Spring (`3`), and Summer (`4`, `5`, and `6`)---within an academic year

For example, for academic year 1995--96, 

- `19951` encodes Fall 95--96 (calendar year 1995)
- `19953` encodes Spring 95--96 (calendar year 1996)
- `19954` encodes Summer 95--96, first session (calendar year 1996)

Different institutions supply data over different time spans. 

```{r}
# Range of data by institution
x <- term[, .(min_term = min(term), max_term = max(term)), by = "institution"]
x[order(min_term, max_term)]
```

Programs are encoded in the `cip6` variable, a 6-digit character based on the 2010 Classification of Instructional Programs (CIP) [@NCES:2010].

```{r}
# A sample of cip6 values
sort(unique(sample(term$cip6, 8)))
```

Student level is used when determining timely completion terms of transfer students.

```{r}
# Possible values
sort(unique(term$level))
```



## `course`

Student-level course information for `r length(unique(course$mcid))` undergraduates including course name and number, credit hours, and student grades for individual students in each term and course. 

The `course` data table has `r nrow(course)` observations of `r ncol(course)` variables structured in block-record form, that is, records associated with a particular ID can span multiple rows---one record for each student in a course in a term. The data dictionary is documented at `?course`.

```{r}
course
```

Sample of the course names. The course names do not contribute to the composite key, thus NAs are possible. 

```{r}
# Sample course names
sample(course$course, 10)
```

Possible values of the faculty rank variable. 

```{r}
# Possible values
sort(unique(course$faculty_rank))
```

Possible values of the type of course.

```{r}
# Possible values
sort(unique(course$type))
```

Sample of the possible values of the MIDFIELD-assigned discipline. 

```{r}
# Sample of possible values
sort(sample(unique(course$discipline_midfield), 20))
```

Using the course abbreviation and number keys, we can count the number of unique courses in these data.  

```{r}
# Unique courses
x <- unique(course[, .(abbrev, number, institution)])
x

# N courses by institution
x <- x[, .N, by = c("institution")]
x[order(-N)]
```




## `degree`

Student-level degree information for `r length(unique(degree$mcid))` undergraduates including institution, program, term, and baccalaureate degree for each student. 

The `degree` data table has `r nrow(degree)` observations of `r ncol(degree)` variables structured in block-record form, that is, records associated with a particular ID can span multiple rows---one record for each student completing a program in a term. Multiple degrees can occur in the same term or in different terms. The data dictionary is documented at `?degree`.

```{r}
degree
```

This dataset contains records for graduates only, thus `degree` has fewer rows that `student`. The `term_degree` variable indicates when the student graduated and the and `cip6` indicates from which program. 

Sample of the degree names. 

```{r}
sort(sample(degree$degree, 10))
```

The number of unique degrees in these data by program code.

```{r}
# Count the number of unique program codes
length(unique(degree$cip6))
```

Number of degrees earned per student. 

```{r}
# Degree count by student
N_degrees <- degree[, .(degree_count = .N), by = "mcid"]

# Number of students by the degree count
N_degrees[, .(N_students = .N), by = "degree_count"]
```

In these practice data, no one earned more than two degrees; 122 students earned two degrees.


```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
