---
title: "Data structure"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})

# Function returning column classes table
library(gt)
var_class <- function(df, caption) {
  x <- t(as.data.frame(lapply(df, 
                              function(x) paste(class(x), 
                                                collapse = ','))))
  x <- data.table(variable = row.names(x), class = x[, 1])
  x |>
    gt() |>
    tab_caption(caption) |>
    tab_options(table.font.size = "small") |>
    opt_stylize(style = 1, color = "gray") |>
    tab_style(
      style = list(cell_fill(color = "#c7eae5")),
      locations = cells_column_labels(columns = everything())
    )
}

library(wrapr, warn.conflicts = FALSE)
library(midfielddata)
suppressPackageStartupMessages(library(data.table))

# data.table printout
options(datatable.print.nrows = 10,
        datatable.print.topn = 3,
        datatable.print.class = TRUE)
```

In this article, we examine the general structure and identify the primary and composite keys for the four data  tables: `student`, `course`, `term`, and `degree`. 



## Load data

*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:

```{r}
#| eval: false
library(wrapr)
library(midfielddata)
library(data.table)
```


*Load data tables.*  &nbsp; Data tables can be loaded individually or collectively as needed.  View data dictionaries via `?student`, `?course`, `?term`, or `?degree`. 

```{r}
# Load one table as needed
data(student)

# Or load multiple tables
data(course, term, degree)
```






## `student`

Data table with `r ncol(student)` variables and `r nrow(student)` rows. The variables are listed in Table 1. 

```{r}
#| echo: false
var_class(student, caption = "Table 1: Variables in student")
```

The primary key is the student ID variable, `mcid`. We can use a function from the wrapr package to check that this column provides a unique key and verify that the number of NA values in the key column is zero. 

```{r}
# Check that the indicated column forms a primary key
DT <- copy(student)
key_col_names <- c("mcid")
wrapr::checkColsFormUniqueKeys(DT, key_col_names)

# Primary key variable should have no missing values
sum(is.na(student$mcid))
```

Rows contain one observation per student.

```{r}
# First few rows and last few rows
student
```

Student IDs and institution names have been anonymized to remove identifiable information. 

```{r}
# Anonymized IDs
sample(student$mcid, 8)

# Anonymized institutions
sort(unique(student$institution))
```

Race/ethnicity and sex are often used as grouping variables.

```{r}
# Possible values
sort(unique(student$race))

# Possible values
sort(unique(student$sex))
```

Counts in each category. 

```{r}
# N by institution
student[order(institution), .N, by = "institution"]

# N by race
student[order(race), .N, by = "race"]

# N by sex
student[order(sex), .N, by = "sex"]
```





## `course`

Data table with `r ncol(course)` variables and `r nrow(course)` rows. The variables  are listed in Table 2.

```{r, comment = ""}
#| echo: false
var_class(course, caption = "Table 2: Variables in course")
```

The key for the course data is a composite key of four variables: student ID, term, course abbreviation, and course number. 

```{r}
# Check that the indicated column forms a primary key
DT <- copy(course)
key_col_names <- c("mcid", "term_course", "abbrev", "number")
wrapr::checkColsFormUniqueKeys(DT, key_col_names)
 
# Composite key variables should have no missing values
sum(is.na(course$mcid))
sum(is.na(course$term_course))
sum(is.na(course$abbrev))
sum(is.na(course$number))
```

Rows contain one observation per student per term per course.

```{r}
# First few rows and last few rows
course
```








## `term`

Data table with `r ncol(term)` variables and `r nrow(term)` rows. The variables   are listed in Table 3.

```{r, comment = ""}
#| echo: false
var_class(term, caption = "Table 3: Variables in term")
```

The key for the term data is a composite key of two variables: student ID and term. 

```{r}
# Check that the indicated column forms a primary key
DT <- copy(term)
key_col_names <- c("mcid", "term")
wrapr::checkColsFormUniqueKeys(DT, key_col_names)

# Composite key variables should have no missing values
sum(is.na(term$mcid))
sum(is.na(term$term))
```

Rows contain one observation per student per term.

```{r}
# First few rows and last few rows
term
```

Terms are encoded `YYYYT`, where 

- `YYYY` is the year at the start of the academic year, and  
- `T` encodes the semester or quarter---Fall (`1`), Winter (`2`), Spring (`3`), and Summer (`4`, `5`, and `6`)---within an academic year

For example, for academic year 1995--96, 

- `19951` encodes Fall 95--96
- `19953` encodes Spring 95--96
- `19954` encodes Summer 95--96 (first session)

Different institutions supply data over different time spans. 

```{r}
# Range of data by institution
term[, .(min_term = min(term), max_term = max(term)), by = "institution"]
```

Programs are encoded in the `cip6` variable, a 6-digit character based on the 2010 Classification of Instructional Programs (CIP) [@cip2010].

```{r}
# A sample of cip6 values
sort(unique(sample(term$cip6, 8)))
```

Student level is used when determining timely completion terms of transfer students.

```{r}
# Possible values
sort(unique(term$level))
```






## `degree`

Data table with `r ncol(degree)` variables and `r nrow(degree)` rows. The variables  are listed in Table 4.

```{r, comment = ""}
#| echo: false
var_class(degree, caption = "Table 4: Variables in degree")
```

The key for the degree data is a composite key of two variables: student ID and degree 

```{r}
# Check that the indicated column forms a primary key
DT <- copy(degree)
key_col_names <- c("mcid", "degree")
wrapr::checkColsFormUniqueKeys(DT, key_col_names)

# Composite key variables should have no missing values
sum(is.na(degree$mcid))
sum(is.na(degree$degree))
```

Rows contain one observation per student per degree.

```{r}
# First few rows and last few rows
degree
```

This dataset contains records for graduates only, thus the number of observations in `degree` (`r nrow(degree)`) is less than the number of observations in `student` (`r nrow(student)`). The `term_degree` and `cip6` variables indicate when and from which program a student graduates.  

Number of degrees earned per student. In these practice data, only one student earned more than one degree. 

```{r}
# Count number of degrees by student
by_id <- degree[, .(degree_count = .N), by = "mcid"]

# Count number of students by number of degrees earned
by_id[, .(N_students = .N), by = "degree_count"]
```


## References
