---
title: "Data linked by student ID"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", 
                              criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

To examine the variables and some representative values in midfielddata, we take a closer look at the records of individual students across the four data tables. 




## Load data

*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfielddata)
library(data.table)

# data.table printout option
options(datatable.print.nrows = 70)
```


*Load data tables.*  &nbsp; Data tables can be loaded individually or collectively as needed. 

```{r}
# Load one table as needed
data(student)

# Or load multiple tables
data(course, term, degree)
```

View data dictionaries on the help pages using `?student`, `?course`, `?term`, `?degree`. 

## Student 1: Degree in psychology and sociology 

We display the records for one specific student, using their ID to subset each data table.

```{r}
# One student ID
id_we_want <- "MCID3112192438"
```





*student.* &nbsp; Yields one row per student.

```{r}
# Observations for a selected ID
DT <- student[mcid == id_we_want]

# display
DT
```







*course.* &nbsp; For this student, the course records span `r nrow(course[mcid == id_we_want])` rows, one row per course.

```{r}
# Observations for a selected ID
DT <- course[mcid == id_we_want]

# display
DT
```

If we omit several columns and rename others, we obtain a more readable output. 

```{r}
# columns to keep
cols_we_want <- c("term_course", "course", "abbrev", "number", "hours_course", "grade")

# subset
DT <- course[mcid == id_we_want, ..cols_we_want]

# order rows
DT <- DT[order(term_course, number)]

# merge course abbreviation and number, shorten some column names
DT <- DT[, .(term = term_course, course, number = paste0(abbrev, "-", number), cr_hr = hours_course, grade)]

# display
DT
```







*Term.* &nbsp; Here, the records span `r nrow(term[mcid == id_we_want])` rows, one row per term. Again, we can do some editing to improve readability, assuming these are variables we might need in an analysis. 

```{r}
# columns to keep
cols_we_want <- c("term", "cip6", "level", "hours_term", "gpa_term", "gpa_cumul")

# subset the data
DT <- term[mcid == id_we_want, ..cols_we_want]

# order rows
DT <- DT[order(term)]

# display
DT
```





*Degree.* &nbsp; In this example, the records span `r nrow(degree[mcid == id_we_want])` rows, one row per degree. The degrees were earned in the same term, Spring 2009.

```{r}
# Observations for a selected ID, showing selected columns
DT <- degree[mcid == id_we_want, .(term_degree, cip6, degree)]

# display
DT
```









## Student 2: Degree in biological sciences


```{r}
# One student 
id_we_want <- "MCID3111315508"
```


*Degree.* &nbsp; In this example, a student has earned two degrees at separate times. Usually we are interested in student records up to the time of their first degree, ignoring subsequent degrees (unless those degrees are relevant to one's study). 

```{r}
# Observations for a selected ID, showing selected columns
DT <- degree[mcid == id_we_want, .(term_degree, cip6, degree)]

# display
DT
```

Thus, when extracting course and term information, we filter for terms no later than the first degree term---in this case, term 19961.

```{r}
# extract the term of the first degree
first_degree_term <- degree[mcid == id_we_want, min(term_degree)]

# display
first_degree_term
```




*student.* &nbsp; As expected, yields one row.

```{r}
# Observations for a selected ID
DT <- student[mcid == id_we_want]

# display
DT
```

*course.* &nbsp; For this student, the records show `r nrow(course[mcid == id_we_want & term_course <= first_degree_term])` courses (one row per course) leading up to thier first degree.

```{r}
# columns to keep
cols_we_want <- c("term_course", "course", "abbrev", "number", "hours_course", "grade")

# subset
DT <- course[mcid == id_we_want, ..cols_we_want]

# retain terms leading up to the first degree
DT <- DT[term_course <= first_degree_term]

# order rows
DT <- DT[order(term_course, number)]

# merge course abbreviation and number, shorten some column names
DT <- DT[, .(term = term_course, course, number = paste0(abbrev, "-", number), cr_hr = hours_course, grade)]

# display
DT
```







*Term.* &nbsp; Here, the records show `r nrow(term[mcid == id_we_want & term <= first_degree_term])` terms (one row per term) leading up to their first degree.

```{r}
# columns to keep
cols_we_want <- c("term", "cip6", "level", "hours_term", "gpa_term", "gpa_cumul")

# subset the data
DT <- term[mcid == id_we_want, ..cols_we_want]

# retain terms leading up to the first degree
DT <- DT[term <= first_degree_term]

# order rows
DT <- DT[order(term)]

# display
DT
```






## Wrap-up

```{r}
# Reset data.table print option
options(datatable.print.nrows = 10)
```





```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
